Type theories as quotient inductive-recursive types

### 2025-02-06 tex/main.tex
Draft S4.2 on the standard model

\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
\usepackage{manfnt}
\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},#1]{#2}}
\newcommand{\Gene}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
\newcommand{\Fredrik}[2][]{\todo[inline,author={Fred},caption={},#1]{#2}}

@@

Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$, so we have the empty substitution $\emptysub$ and substitution extensions $\sigma, t$ by some term~$t$:
\begin{alignat*}{3}
  \emptysub & : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A),
\end{alignat*}
Note that type substitution $[\sigma]\;A$ is needed, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
As substitution calculus is the initial cwf with a constant type $\UU$, substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$
\begin{alignat*}{3}
  \idS & : \implicit{\Gamma}\;\Sub\;\Gamma\;\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta
\end{alignat*}
satisfying certain laws (introduced later).
Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$
for its \emph{tail} and its \emph{head}:
\begin{alignat*}{3}
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\,\sigma ]\; A)
\end{alignat*}
respectively.
Again, type substitution $[\pi_1\,\sigma]\,A$ is needed for~$\pi_2$, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
Finally, we have term substitution:
\begin{alignat*}{3}
  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\; A).
\end{alignat*}
such that $[\sigma]\,t$ is of $\Tm\;\Gamma\;([\sigma]\,A)$.
Note that $[\blank]\blank$ is overloaded for type and term substitution.

@@

One possibility is to define substitution calculus as a QII`R' type where type substitution, as a simultaneously defined recursion, does compute.
Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may define it by
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU & = \UU
\end{alignat*}
Note that this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneously defined function does not refer to any inductive type being defined.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.}
Still, although $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ can be proved equal to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, they are not definitional.
This definition fails to remove any transport in the definition of substitution calculus.

Our second attempt is to define type substitution with additional rules:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally.
By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching.
\LT[noinline]{How about \Coq and \Lean?}
In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
\end{alignat*}
where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
%As a result, neither  are definitional.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \reduce \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \reduce \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \reduce \UU,
\]
making $[\sigma]\;\UU = \UU$ propositional instead.
Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
That is, using the standard semantics of function definition, we are not able to make type substitution definitional for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$.

To ensure that type substitution is definitional for all \emph{rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as it is locally confluent and terminating.
We end up with the following definition for type substitution:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is confluent.\footnote{%
\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check}.
This option is turned on globally in our formal implementation.}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

@@

With this definition of type substitution, other equality constructors can be introduced without any $\transp$, we list updated constructors as follows.
\begin{alignat*}{5}
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} &&\sigma ; (\tau , t) && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau), [ \sigma ]\;t \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\, A} &&  t \\
  [\idS]t         & : \implicit{\Gamma, A, t} && {[\idS]\,t}          && =^{\Tm\,\Gamma\,A} && t \\
  [;]t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\sigma ; \tau]\,t} && =^{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\;[ \tau ]\;t}
\end{alignat*}

@@

\section{The elimination principle and models} \label{sec:models}
\LT{Terminology: the eliminator principle vs.\ eliminator vs.\ induction principle?}
\LT{%
\begin{enumerate}
  \item Show that two definitions are equivalent (ongoing).
  \item Show that methods are equivalent (ongoing).
\end{enumerate}
}
With the QIIR definition of a type theory, we show how to \emph{eliminate} $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}) .
Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and also show that our definition is actually equivalent to its QII counterpart syntactically and semantically (\cref{subsec:equivalence}).

\subsection{The elimination principle} \label{subsec:eliminator}
To eliminate $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$, we have to specify types
\begin{align*}
  \Ctx^M & : \Ctx \to \Set &
  \Sub^M & : \implicit{\Gamma, \Delta}\,\Ctx^M\;\Gamma \to \Ctx^M\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
  \Ty^M  & : \implicit{\Gamma}\,\Ctx^M\;\Gamma \to \Ty\;\Gamma \to \Set & 
  \Tm^M  & : \implicit{\Gamma, A}\; (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Tm\;\Gamma\;A \to \Set
\end{align*}
called the \emph{motivates} as the target types for the elimination principle:
\begin{alignat*}{5}
  \elim_{\Ctx} & : (\Gamma : \Ctx)   && \to \Ctx^M\;\Gamma \\
  \elim_{\Ty}  & : (A : \Ty\;\Gamma) && \to \Ty^M\;(\elim_{\Ctx}\,\Gamma)\;A \\
  \elim_{\Sub} & : (\sigma : \Sub\;\Gamma\;\Delta) && \to \Sub^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ctx}\;\Delta)\;\sigma \\
  \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) && \to \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;t.
\end{alignat*}

As inhabitants of these inductive types are formed by constructors, each constructor $c$ is interpreted by a corresponding functions $c^M$ indexed by $c$ such as
\begin{alignat*}{3}
  \emptyctx_{\Ctx}^M     & :                      && \Ctx^M\;\emptyctx \\
  \blank,_{\Ctx}^M\blank & : \implicit{\dots}\; && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
  \emptysub_{\Sub}^M     & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
  \blank,_{\Sub}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
  \idS^M          & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
  \blank;\blank   & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Sub^M\;\Delta^M\;\Theta^M\;\tau \to \Sub^M\;\Gamma^M\;\Theta^M\;(\sigma;\tau)\\
  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t)
\end{alignat*}
and simultaneously defined functions are also given by corresponding functions 
\begin{alignat*}{3}
  [\blank]_{\Ty}^M\blank & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t)
\end{alignat*}
subject to the corresponding equations \emph{and function clauses}, e.g., 
\begin{alignat*}{5}
  [\idS]_{\Ty}^M & : \implicit{\dots} && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M \\
  [;]_{\Ty}^M & : \implicit{\dots} && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
  [\idS]_{\Tm}^M & : \implicit{\dots} && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS^M]}} && t^M
\end{alignat*}
called \emph{methods}.
For brevity, we omit the rest of methods and refer to the Agda implementation for their complete definition.
\LT[noinline]{Add a link.}

\begin{definition}
  An \emph{eliminator} for the QIIR definition of type theory consists of motives and methods above.
\end{definition}

@@

\subsection{The standard model} \label{subsec:std-model}
Our QIIR definition of type theory has a standard model by giving the following motives and methods.

\subsection{The equivalence} \label{subsec:equivalence}
\section{Formal implementation}\label{sec:formal-implementation}
\LT{Mention the NbE for SC here}
\LT{Compare some formal proofs with pen-and-paper proofs here}
\section{Discussion}
\subsection{Related work}
\paragraph*{Formalisation of type theory in type theory}
\cite{Danielsson2006,Altenkirch2016a,Chapman2009}
\cite{Altenkirch2017}
\cite{Munoz1998}
\cite{Dybjer1996,Castellan2021}
\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}

>>>>

\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
\usepackage{manfnt}
\usepackage{underscore}
\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},#1]{#2}}
\newcommand{\Gene}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
\newcommand{\Fredrik}[2][]{\todo[inline,author={Fred},caption={},#1]{#2}}

@@

Substitutions from $\Gamma$ to $\Delta$ can be understood intuitively as lists of terms of type $A$ under the context $\Gamma$ for each $A$ in $\Delta$, so we have the empty substitution $\emptysub$ and substitution extensions $\sigma, t$ by some term~$t$:
\begin{alignat*}{3}
  \emptysub & : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx \\
  \blank,\blank & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;([ \sigma ]\, A) \to \Sub\;\Gamma\;(\Delta, A),
\end{alignat*}
Note that type substitution $[\sigma]\;A$ is needed, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
Substitutions also have the identity substitution $\idS$ and composition $\blank;\blank$ (because substitution calculus is the initial \emph{category} with families), 
\begin{alignat*}{3}
  \idS & : \implicit{\Gamma}\;\Sub\;\Gamma\;\Gamma \\
  \blank;\blank & : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta
\end{alignat*}
satisfying certain laws (introduced later as equality constructors).
Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$
for its \emph{tail} and its \emph{head}:
\begin{alignat*}{3}
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;([ \pi_1\,\sigma ]\; A)
\end{alignat*}
respectively.
Again, type substitution $[\pi_1\,\sigma]\,A$ is needed for~$\pi_2$, because $A$ is well-formed under the context $\Delta$ instead of $\Gamma$.
Finally, we have term substitution:
\begin{alignat*}{3}
  [\blank] \blank & : \implicit{\Gamma,\Delta, A}\;(\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Delta\;A \to \Tm\;\Gamma\;([\sigma]\; A).
\end{alignat*}
such that $[\sigma]\,t$ is of $\Tm\;\Gamma\;([\sigma]\,A)$.
Note that $[\blank]\blank$ is overloaded for type and term substitution.

@@

One possibility is to define substitution calculus as a QII`R' type where type substitution, as a simultaneously defined recursion, does compute.
Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU & = \UU
\end{alignat*}
Note that this definition is \emph{not} inductive-recursive in the sense of Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}: in their schema, the codomain of a simultaneously defined function does not refer to any inductive type being defined.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisation of type theory~\cite{Danielsson2006}.}
Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
This definition fails to remove any transport in the definition of substitution calculus.

Our second attempt is to define type substitution with additional rules:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} definitionally.
By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching.
\LT[noinline]{How about \Coq and \Lean?}
In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
\begin{alignat*}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
\end{alignat*}
where $\reduce$ is used to emphasise the reduction (which entails the definitional equality): we have done nothing different from our first attempt! 
%As a result, neither  are definitional.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} definitional, but then \eqref{eq:type-sub-at2-1} will be translated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \reduce \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \reduce \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \reduce \UU,
\]
making $[\sigma]\;\UU = \UU$ propositional instead.
Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
That is, using the standard semantics of function definition, we are not able to make type substitution definitional for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

To ensure that type substitution is definitional for all \emph{rules}, we turn to a definition by overlapping patterns (\Cref{sec:meta:overlapping}) instead to make each clause definitional. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as they are locally confluent and terminating.
We end up with the following definition for type substitution:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is confluent.\footnote{%
\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check}.
This option is turned on globally in our formal implementation.}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

@@

With this definition of type substitution, other equality constructors can be introduced without any $\transp$, updated constructors are listed below.
\begin{alignat*}{5}
  \mathsf{concat} & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, A, t} &&\sigma ; (\tau , t) && =^{\Sub\,\Gamma\,(\Theta, A)} &&  (\sigma ; \tau), [ \sigma ]\;t \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\, A} &&  t \\
  [\idS]t         & : \implicit{\Gamma, A, t} && {[\idS]\,t}          && =^{\Tm\,\Gamma\,A} && t \\
  [;]t            & : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t} && {[\sigma ; \tau]\,t} && =^{\Tm\,\Gamma\,[\sigma ; \tau] A} && {[ \sigma ]\;[ \tau ]\;t}
\end{alignat*}

@@

\section{The elimination principle and models} \label{sec:models}
\LT{Terminology: the eliminator principle vs.\ eliminator vs.\ induction principle?}
\LT{%
\begin{enumerate}
  \item Show that two definitions are equivalent (ongoing).
  \item Show that methods are equivalent (ongoing).
\end{enumerate}
}
With the QIIR definition of a type theory, we show how to \emph{eliminate} $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}) .
Then, we can give this QIIR definition of type theory a \emph{standard model} (\cref{subsec:std-model}) and compare our QIIR definition with its QII counterpart syntactically and semantically (\cref{subsec:equivalence}).

\subsection{The elimination principle} \label{subsec:eliminator}
\LT{We shall say something about the elimination/induction principle in \cref{sec:meta}}

To eliminate $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$, we have to specify types
\begin{align*}
  \Ctx^M & : \Ctx \to \Set &
  \Sub^M & : \implicit{\Gamma, \Delta}\,\Ctx^M\;\Gamma \to \Ctx^M\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
  \Ty^M  & : \implicit{\Gamma}\,\Ctx^M\;\Gamma \to \Ty\;\Gamma \to \Set & 
  \Tm^M  & : \implicit{\Gamma, A}\; (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Tm\;\Gamma\;A \to \Set
\end{align*}
called the \emph{motivates} as the target types for the elimination:
\begin{alignat*}{5}
  \elim_{\Ctx} & : (\Gamma : \Ctx)   && \to \Ctx^M\;\Gamma \\
  \elim_{\Ty}  & : (A : \Ty\;\Gamma) && \to \Ty^M\;(\elim_{\Ctx}\,\Gamma)\;A \\
  \elim_{\Sub} & : (\sigma : \Sub\;\Gamma\;\Delta) && \to \Sub^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ctx}\;\Delta)\;\sigma \\
  \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) && \to \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;t.
\end{alignat*}
Each constructor $c$ is interpreted by a corresponding functions $c^M$ indexed by $c$, i.e.\ 
\begin{alignat*}{3}
  \emptyctx_{\Ctx}^M     & :                      && \Ctx^M\;\emptyctx \\
  \blank,_{\Ctx}^M\blank & : \implicit{\dots} && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
  \UU^M & : \implicit{\dots} && \Ty^M\;\Gamma^M\;\UU \\
  \emptysub_{\Sub}^M     & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
  \blank,_{\Sub}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
  \idS^M          & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t) \\
                   && \vdots
\end{alignat*}
and simultaneously defined functions are also given by corresponding functions 
\begin{alignat*}{3}
  [\blank]_{\Ty}^M\blank & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) \\
                         && \vdots
\end{alignat*}
subject to the corresponding equations \emph{and function clauses}, e.g., 
\begin{alignat*}{5}
  [\idS]_{\Ty}^M & : \implicit{\dots} && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M \\
  [;]_{\Ty}^M & : \implicit{\dots} && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
  [\idS]_{\Tm}^M & : \implicit{\dots} && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS^M]}} && t^M \\
                 && \vdots
\end{alignat*}
called \emph{methods}.
For brevity, we omit the rest of methods and refer to the Agda implementation for them.
\LT[noinline]{Add a link.}
We also have omitted the list of implicitly quantified variables: each variable $A$ implicitly quantified in a constructor $c$ has accordingly an $A$-indexed variable $A^M$ in $c^M$.

\begin{definition}
  An \emph{eliminator} for the QIIR definition of type theory is a tuple of motives and methods above.
\end{definition}

@@

\subsection{The standard model} \label{subsec:std-model}
